---
title: "Week 3 Analytical Assigment"
author: "M. Onimus"
date: "03/25/2020"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float:
      collapse: yes
      smooth_scrolling: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(haven)
library(car)
library(broom)
library(oddsratio)
library(jtools)
library(kableExtra)
library(nnet)
library(MASS)

```

# R Set up

Before getting started, I have loaded in a few packages to help us with the analysis.

-   `tidyverse`: A collection of r packages used to make data import, manipulation, and analysis easier.
-   `here`: A package that makes it easier to navigate around the folder structure.
-   `haven`: A package with a set of functions for reading in SAS files.
-   `jtools`: A package for making publication ready tables.
-   `kableExtra`: A package for printing pretty tables to html.
-   `nnet`: A package with a multinomal regression function
-   `MASS`: A package with an ordinal regression function

# From the Assignment

## Introduction

For the multinomial assignment you will use the data set log.disease_type, which has the following variables:

•	subject: observation identifier
•	AGE: patient age in years
•	SEX: patient gender with values of 'F' and 'M'
•	Specialist_6mos: patient seen by specialist in previous 6 months
•	newly_diagnosed: patient was diagnosed with disease in past 12 months 1=yes 0=no
•	chronic: family of chronic illness - 0=Respiratory; 1=Cardiac; 2=Circulatory
•	health_rating: self-rated health 0=poor; 1=average; 2=good 
•	age_cat: categorical age variable with 10 categories each with 10-year increments from 0-90+ 
•	age_cat2: categorical age variable 5 categories of different bin sizes
•	age_0_9 to age_90: 10 dummy variables for age_cat categories 
•	age_0_17 to age_65: 5 dummy variables for age_cat2 categories 

# Multinomial Logistic Regression

## Read the data in

In order to read the data into R, we will need to use the `haven` package.

```{r readData}

multi <- read_sas(here("data/disease_type.sas7bdat"))


```

This data set contains `r NROW(multi)` rows with `r NCOL(multi)` variables, as described in the data overview.

## Questions 1-2

Run the PROC LOGISTIC code below and to produce output for a multinomial logistic regression model. The model can be used to describe characteristics of people with different chronic diseases using Circulatory disease (ref='2') as the reference group. 

We will need to relevel the data and then build the appropiate model.

```{r multi1}

multi <- multi %>% 
  mutate(SEX = as.factor(SEX),
         newly_diagnosed = as.factor(newly_diagnosed),
         chronic = as.factor(trimws(chronic)),
         chronic = relevel(chronic, ref = "2"),
         SEX = relevel(SEX, ref = '1'),
         newly_diagnosed = relevel(newly_diagnosed, ref = '1'))

# just to check the levels
# levels(multi$chronic)
# levels(multi$SEX)
# levels(multi$newly_diagnosed)

model1 <- nnet::multinom(chronic ~ AGE + SEX + newly_diagnosed,
                         data = multi)


```

Here is the summary of the model:

```{r mod1}

model1

```

Here is a summary of odds ratio estimates:

```{r odds1}

exp(coef(model1))


```

Interpret the odds ratios for age, sex, and newly_diagnosed.

NEED TO DO THIS!!


## Questions 3-4

Run a model using Respiratory as the reference group.

```{r multi2}

multi2 <- multi %>% 
  mutate(SEX = as.factor(SEX),
         newly_diagnosed = as.factor(newly_diagnosed),
         chronic = as.factor(trimws(chronic)),
         chronic = relevel(chronic, ref = "0"),
         SEX = relevel(SEX, ref = '1'),
         newly_diagnosed = relevel(newly_diagnosed, ref = '1'))

model2 <- nnet::multinom(chronic ~ AGE + SEX + newly_diagnosed,
                         data = multi2)


```

Here is the summary of the model:

```{r mod2}

model2

```

Here is a summary of odds ratio estimates:

```{r odds2}

exp(coef(model2))


```

Interpret the odds ratios for newly_diagnosed.

*Need to do this*

## Question 5

What information was gained by changing the reference group? That is, what do we know from the results of the new model that we did not know from the previous model?

*Need to do this*


## Questions 6-7

Experiment with adding new variables to the model (listed above) to find a model that has fits the data better/has better explanatory power that the model with age, sex, and newly_diagnosed

```{r multi3}

multi3 <- multi %>% 
  mutate(SEX = as.factor(SEX),
         newly_diagnosed = as.factor(newly_diagnosed),
         health_rating = as.factor(trimws(health_rating)),
         age_cat = as.factor(age_cat),
         age_cat2 = as.factor(age_cat2),
         Specialist_6mos = as.factor(Specialist_6mos),
         chronic = as.factor(trimws(chronic)),
         chronic = relevel(chronic, ref = "0"),
         SEX = relevel(SEX, ref = '1'),
         newly_diagnosed = relevel(newly_diagnosed, ref = '1'))

model3 <- nnet::multinom(chronic ~ age_cat2 + SEX + newly_diagnosed + health_rating + Specialist_6mos,
                         data = multi3)


```

Here is the summary of the model:

```{r mod3}

model3

```

Here is a summary of odds ratio estimates:

```{r odds3}

exp(coef(model3))

```


The 3rd model I fitted has an AIC of 31714.16 which is less than both the previous fitted models.


# Ordinal Logistic Regression

## Read the data in

The data set log.cardiac_males is a subset of the log.disease_type data set containing only males with cardiac disease.

In order to read the data into R, we will need to use the `haven` package.

```{r readData2}

ord <- read_sas(here("data/cardiac_males.sas7bdat")) %>% 
  mutate(SEX = as.factor(SEX),
         newly_diagnosed = as.factor(newly_diagnosed),
         health_rating = as.factor(trimws(health_rating)),
         age_cat = as.factor(age_cat),
         age_cat2 = as.factor(age_cat2),
         Specialist_6mos = as.factor(Specialist_6mos),
         Specialist_6mos = relevel(Specialist_6mos, ref = '1'),
         newly_diagnosed = relevel(newly_diagnosed, ref = '1'))


```

This data set contains `r NROW(data)` rows with `r NCOL(data)` variables, as described in the data overview.

## Questions 1-4

Run the ordinal model using the code below

```{r}

model4 <- polr(health_rating ~ newly_diagnosed + Specialist_6mos,
               data = ord, method = 'logistic', Hess = TRUE)

#model4



```

Here is the summary of the model:

```{r}

summary(model4)

```

Here is a summary of odds ratio estimates:

```{r}

exp(coef(model4))

```

Yes, an ordinal model is reasonable for this data since we are measuring health rating which has an order to it. 

The model fits the data reasonably well with an AIC of 963 and a residual deviance of 955.

The odds ratio for a patient with newly diagnosed disease in the last 12 months to have worse health (1 or 2) is 1.55 times compared to a patient without a new diagnosis.  

## Questions 5-6

**NOTE: I did not look into how to execute this statistics in R.  I have used SAS results for the below answers.**

The predicted probability of a male with cardiac disease (everyone in this dataset) rating their health as Poor if they have been newly diagnosed and have seen a specialist in the past 6 months is 39.7%

The predicted probability of having self-rated health of Poor or Average for someone who is not newly diagnosed and has not seen a specialist is 31.9%.